## Basics:

plyr and dplyr don't "add" behaviour to data frames, so we still use them (mostly):

```{r}
myDT <- data.frame(
    number=1:3,
    letter=c('a','b','c')
  ) # like data.frame constructor
myDT
````

We still rely on basics for reading data from CSVs and the like:

```{r}
D <- read.csv("TB_Burden_Data.csv")
head(names(D))
````

##plyr is bare-bones:

It doesn't give us the nice extraction syntax data.table provides, so there's still some repetition:

```{r}
mean(D[D$country=='Afghanistan','e_prev_100k'])
```
However, we can improve on that slightly using `with`:

```{r}
with(D, mean(e_prev_100k[country=='Afghanistan']))
```

We extract multiple columns with base R:
```{r}
head(D[,c('e_prev_100k','e_prev_100k_lo',
    'e_prev_100k_hi')])
```

Summarisation is similar to data.table, however there's no need to use `list` on the new columns, and we specify the group before the application.

```{r}
library(plyr)
E <- D[with(D, country=='Afghanistan'),] 
ddply(E, .(), summarise, mid=mean(e_prev_100k),
       lo=mean(e_prev_100k_lo),
       hi=mean(e_prev_100k_hi) )
```

`plyr` functions expect groupings, which is why the `.()` is needed. plyr is designed for the split-apply-combine pipeline.

Means for each country; take `D`, group by `country`, calculating mid. Summarise tells plyr to use one value for each group, and throw away all the other columns.
```{r}
head(ddply(D, .(country), summarise, mid=mean(e_prev_100k)))
```

Pete's more complicated example:

```{r}
head(ddply(D, .(country, year > 2000), summarise, lo=mean(e_prev_100k_lo),
       hi=mean(e_prev_100k_hi)))
```

Sorting is a little bit nicer than base R:

```{r}
plot(arrange(D, e_prev_100k)$e_prev_100k)
```

Adding new columns to a data frame:
```{r}
head(mutate(D, country_t = paste0(country,year)))
```

Most of the gotchas from data table don't exist:

 * assignment copies by value
 * column selection works as normal
 
all of the gotchas of data frames persist, however.

Bonus: read all CSV files matching "test*.csv" and produce one data frame back:
```{r}
ldply(list.files(pattern="test.*\\.csv"), read.csv)
````
Same again, but including filename:
```{r}
ldply(list.files(pattern="test.*\\.csv"), function(file){ df <- read.csv(file); df$name <- file; df})
````
This is where plyr can be very useful; `ply` commands can be nested, and you can write functions to apply and let R deal with combining the result. Easier than loops  using rbind, and usually faster.

Plyr can parallelise by setting `.parallel` to `TRUE`, and can show progress bars for long-running operations.